Evolution of Processes in Delivering Reliable Software in the European Scientific Arena

### 2018-06-11 paper-ieee-trans.tex
MI/5 page 6 from reviewer#1: devops CD rephrase and referencing

\usepackage{graphicx}
\usepackage{array}
\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage{url}
\usepackage{cite}
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % \DeclareGraphicsExtensions{.eps}
\fi

%\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
\title{Evolution of Methodologies in Delivering Reliable Software in the European Scientific Arena}
%\title{Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis}
%\title{15 years of European scientific software development: evolution and methodologies}

@@

% The paper headers
%\markboth{IEEEtran for IEEE Journals}

@@

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e--Infrastructure creation, operation and management.
% the creation of software development and e-Infrastructure coordination dedicated projects.
This work examines the evolution of the software engineering methodologies used in past and
present European Commission--funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.
\end{abstract}

@@

\IEEEpeerreviewmaketitle

\section{Introduction}

\IEEEPARstart{D}{uring} the last 15 years the provision of an European e--Infrastructure,
supporting unified access to large--scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan--European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
% Comment from reviewer
%(Update: I see this is discussed later... so it would be good just to signpost at the start that a) the focus is really on "grids", and b) that specific examples will be given.
LHC particle accelerator. A significant number of innovation projects and technology transfer
activities -- later discussed in Section \ref{sec:ev} -- from the European Commission
(EC)-funded 5th, 6th and 7th Framework Programmes, and recently from the Horizon 2020 programme
\cite{h2020} had as their main goal to develop software solutions that provide a seamless and
transparent access of researchers to distributed computing and data resources.

% Comment
%%top of second column:  think about saying this more gently
Major developments were carried out throughout the years focused on providing new
functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
The released software was less tested, resulting in a high rate of
rollbacks and patches applied to production systems. Thereby, the stability and
% Comment
%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.
% Comments
%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
It became then apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, incoming projects progressively
devoted a substantial part of the software lifecycle to the improvement of the quality of the
software being delivered. At this point, evolving software engineering practices were considered
and gradually adopted to define and implement quality procedures, the organization of
cross--functional teams and deployment of pilot testbeds.

Analysing the EC--funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. The
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization) motivated this trend. All of which
complemented by the emergence of automation and event--response tools.

@@

The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC--projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section \ref{sec:con}.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way for the following ones, according to a logic evolution in which the role of the quality and reliability of the software have
consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC--projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E--sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI--Inspire &
Integrated Sustainable Pan--European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO--DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

\begin{table*}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force            &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling             &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling              &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies                &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection             &       &       &       &$\surd$&       &$\surd$\\
Build/Testing Management Procedure &$\surd$&$\surd$&$\surd$&$\surd$&       &$\surd$\\
Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition        &       &       &$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
Auto--generated Documentation      &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD) &       &       &       &       &       &$\surd$\\
\hline
\hline
\end{tabular}
%\end{adjustbox}
\end{table*}

\subsection{Enabling Grids for E--sciencE}
% Comment
%%P2 C2 L24: Typographical: It would be good to merge the three citations into one bracket/\cite
The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee,cordis:egee2,cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper, it
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, management of new technologies to put those methodologies
into practice, compilation and analysis of an appropriate set of metrics to measure the
% Comment
%%L26-29 Missing verb
%%(not a sentence—no verb) 
quality of the implemented solutions. All the above with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

@@

%The {\sl INDIGO--DataCloud} project reflects the progress made in software
%quality and reliability aspects throughout the past scientific European
%experiences. Nevertheless (I would remove this), the project's Quality Assurance procedures are also
%highly influenced by the insights of current big worldwide collaborations of
%software development. These collaborations have continuously inspired the use of new
%technologies and procedures to increase the robustness and quality of the
%software being delivered. In this scenario, the implementation of the DevOps culture
%is one of the major outcomes, as this paradigm has been progressively adopted by the
%project.

@@

\item \textit{Unit and Functional testing}.
%% Comment from reviewer       
%% These percentages appear as vague approximations by eye, please add the precise numbers rounded.
Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition and periodic execution
of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's 
internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage 
for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software 
component between the two major releases, denoting an incremental trend over the course of the project: 
code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
the project's first (INDIGO-1) and second (INDIGO-2) releases. By INDIGO-2, 53\% of the software components
(17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation 
(70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.

@@

\item \textit{Code review}.
% Comment
%%Represents …   (not a sentence) 
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human--based
revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
and documentation completeness. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforced,
for any piece of produced software, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline to have a per--change trend evolution of the common source code
related metrics, such as SLOC or the cyclomatic complexity.

@@

% Comment
%%not a sentence, again
Jenkins, the event--response CI system, and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

\subsubsection{Continuous Delivery}
% Comment from reviewer
%Please add a reference.
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually are more correlated. The software updates of
{\sl INDIGO--DataCloud} products, that has been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

@@

% Comment
%%L42-44 not clear, please rephrase
Software lifecycle demands as well from a continuous measurement and analysis
process that permits timely reactions to discover and fix defects throughout the
development, release and post--release or maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

@@

%{\sl DataGrid - Research and Technological Development for an International Data Grid} project
%has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
%FP5-IST - Programme for research, technological development and demonstration on a "User--friendly information society, 1998-2002" (IST--2000--25182E)
%{\sl EGEE - Enabling grids For E--science} project
%has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
% FP6-INFRASTRUCTURES - Research infrastructures: Specific programme for research, technological development and demonstration: "Structuring the European Research Area" under the Sixth Framework Programme 2002-2006, (INFSO--RI--508833)
%{\sl EGEE - Enabling grids For E--science--II} project
% has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
% FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high--capacity and high--speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)
%{\sl EGEE - Enabling grids For E--science--III} project has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
%  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e--Science Grid infrastructures", (INFSO--RI--222667)

@@

@@

\bibitem{cordis:datagrid}
P. Kunszt,
European DataGrid project: status and plans,
\emph{Nucl. Instr. Meth. Phys. Res. A},
vol. 502, n. 2, pp. 376--381, 2003.
%\emph{DataGrid project}, European Community Research and Development Information Service (CORDIS).
%[Online] Available: \url{http://cordis.europa.eu/project/rcn/53665\_en.html}
%[Accessed 4 Jun. 2018].

@@

\bibitem{cordis:egee}
F. Gagliardi and M. E. B\'{e}gin,
EGEE -- providing a production quality grid for e-science,
in \emph{2005 IEEE Inter. Symp. on Mass Storage Systems and Technology},
pp. 88-92, 2005.
%\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community Research and
%Development Information Service (CORDIS). 
%[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}
%[Accessed 4 Jun. 2018].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% TOBE checked if there are papers on these
\bibitem{cordis:egee2}
\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community Research and
Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189\_en.html}
[Accessed 4 Jun. 2018].

\bibitem{cordis:egee3}
\emph{Enabling Grids for E--sciencE--III (EGEE--III)} project, European Community
Research and Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264\_en.html}
[Accessed 4 Jun. 2018].
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% TOBE checked if there are papers on these
%\bibitem{cordis:etics}
%\emph{E--Infrastructure for Testing, Integration and Configuration of Software (ETICS)}
%project, European Community Research and Development Information Service (CORDIS).
%[Online] Available: \url{http://cordis.europa.eu/project/rcn/80138\_en.html}
%[Accessed 4 Jun. 2018].

%\bibitem{cordis:etics2}
%\emph{E--Infrastructure for Testing, Integration and Configuration of Software -
%Phase 2 (ETICS 2)} project, European Community Research and Development
%Information Service (CORDIS).
%[Online]. Available: \url{http://cordis.europa.eu/project/rcn/86604\_en.html}
%[Accessed 4 Jun. 2018].
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\bibitem{cordis:emi}
C. Aiftimiei, et al.,
Towards next generations of software for distributed infrastructures: The European Middleware Initiative,
in \emph{2012 IEEE 8th Inter. Conf. on E-Science},
pp. 1--10, 2012.
%\emph{European Middleware Initiative (EMI)} project, European Community
%Research and Development Information Service (CORDIS).
%[Online] Available: \url{http://cordis.europa.eu/project/rcn/95311\_en.html}
%[Accessed 4 Jun. 2018].

@@

>>>>

\usepackage{graphicx}
\usepackage{array}
\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage{url}
\usepackage{cite}

\begin{document}
\title{Evolution of Processes in Delivering Reliable Software in the European Scientific Arena}

@@

@@

\begin{abstract}

From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e--Infrastructure creation, operation and management.
This work examines the evolution of the software engineering methodologies used in past and
present European Commission--funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.

\end{abstract}

@@

\IEEEpeerreviewmaketitle
\section{Introduction}

\IEEEPARstart{D}{uring} the last 15 years the provision of an European e--Infrastructure,
supporting unified access to large--scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan--European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and technology transfer
activities -- later discussed in Section~\ref{sec:ev} -- from the European Commission
(EC)-funded 5th, 6th and 7th Framework Programmes, and recently from the Horizon 2020 programme
\cite{h2020} had as their main goal to develop software solutions that provide a seamless and
transparent access of researchers to distributed computing and data resources.

Major developments carried throughout the years where focused on providing new
functionalities, while in the early days, the adoption of software engineering methodologies had a lower priority.
The released software was less tested, resulting in a higher rate of
rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
% Comment
%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities \cite{aiftimiei}.
% Comments
%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
It became then apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, incoming projects progressively
devoted a substantial part of the software lifecycle to the improvement of the quality of the
software being delivered. At this point, evolving software engineering practices were considered
and gradually adopted to define and implement quality procedures, the organization of
cross--functional teams and deployment of pilot testbeds.

Analysing the EC--funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures \cite{aiftimiei}. The
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend. All of which
are complemented by the emergence of automation and event--response tools.

@@

The reminder of this paper is
organized as follows. Section~\ref{sec:ev} details the evolution of research on
software reliability in EC--projects. Section~\ref{sec:ntsr} provides
information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.

@@

The EC projects considered in this paper are listed chronologically in Table~\ref{tab:eup}.
Each project paved the way for the following ones, according to a logic evolution in which
the role of the quality and reliability of the software have
consistently increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
committed to and aware of software reliability practices.
Table~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

\begin{table}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{List of EC--projects}
\label{tab:eup}
\centering
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E--sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI--Inspire &
Integrated Sustainable Pan--European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO--DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

\begin{table*}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force            &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling             &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling              &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies                &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection             &       &       &       &$\surd$&       &$\surd$\\
Build/Testing Management Procedure &$\surd$&$\surd$&$\surd$&$\surd$&       &$\surd$\\
Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition        &       &       &$\surd$&$\surd$&$\surd$&$\surd$\\
Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
Auto--generated Documentation      &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD) &       &       &       &       &       &$\surd$\\
\hline
\hline
\end{tabular}
\end{table*}

\subsection{Enabling Grids for E--sciencE}

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee,cordis:egee2,cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities~\cite{egee}. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper, it
leveraged the lessons learned from the previous experiences described in
Section~\ref{sec:ev}. The expertise gathered throughout these years was highly
profitable regarding the evaluation and application of innovative software engineering
methodologies, management of new technologies to put those methodologies
into practice, compilation and analysis of an appropriate set of metrics to measure the
quality of the implemented solutions.
An additional level of complexity, comes from the fact that these large collaborative projects
have an extensive list of partners.

@@

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition and periodic execution
of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's
internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage
for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
component between the two major releases, denoting an incremental trend over the course of the project:
code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
the project's first (INDIGO-1) and second (INDIGO-2) releases. By INDIGO-2, 53\% of the software components
(17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation
(70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.

@@

\item \textit{Code review}.
This phase is the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human--based
revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
and documentation completeness. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}

The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforced,
for any piece of produced software, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline to have a per--change trend evolution of the common source code
related metrics, such as SLOC or the cyclomatic complexity.

@@

Jenkins is an event--response CI and Docker container
provisioning system that provides instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

\subsubsection{Continuous Delivery}
% Comment from reviewer
%Please add a reference.
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the 
development teams to act promptly based on the regular user feedback \cite{chen2015}. The software updates of
{\sl INDIGO--DataCloud} products, that has been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

@@

The implementation of a Software lifecycle phases, implies a continuous measurement and analysis
process that allows timely reactions to discover and fix defects throughout the
development, release and maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

@@

@@

\bibitem{aiftimiei}
A. Aiftimiei, and D. Dongiovanni, et al.,
Improving the quality of EMI Releases by leveraging the
EMI Testing Infrastructure.
\emph{Journal of Physics: Conference Series},
vol. 396, 2012.

@@

\bibitem{soltesz}
S. Soltesz, and H. P\"{o}tzl, et. al.,
Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,
\emph{SIGOPS Oper. Syst. Rev.},
vol. 41, n. 3, pp. 275--287, 2007.

\bibitem{cordis:datagrid}
P. Kunszt,
European DataGrid project: status and plans,
\emph{Nucl. Instr. Meth. Phys. Res. A},
vol. 502, n. 2, pp. 376--381, 2003.

@@

\bibitem{cordis:egee}
\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community Research and
Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}
[Accessed 4 Jun. 2018].

\bibitem{cordis:egee2}
\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community Research and
Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189\_en.html}
[Accessed 4 Jun. 2018].

\bibitem{cordis:egee3}
\emph{Enabling Grids for E--sciencE--III (EGEE--III)} project, European Community
Research and Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264\_en.html}
[Accessed 4 Jun. 2018].

\bibitem{egee}
F. Gagliardi and M. E. B\'{e}gin,
EGEE -- providing a production quality grid for e-science,
in \emph{2005 IEEE Inter. Symp. on Mass Storage Systems and Technology},
pp. 88-92, 2005.

@@

@@

\bibitem{cordis:emi}
C. Aiftimiei, et al.,
Towards next generations of software for distributed infrastructures: The European Middleware Initiative,
in \emph{2012 IEEE 8th Inter. Conf. on E-Science},
pp. 1--10, 2012.

@@

\bibitem{chen2015}
L. Chen,
Continuous delivery: Huge benefits, but challenges too,
\emph{IEEE Software},
vol. 32, no 2, pp. 50--54, 2015.
